---
title: "Introduction to Regression and Least Squares"
author: "Chee Loong"
date: "10/31/2017"
output: html_document
---

***
## Galton's Data

* Let's look at the marginal (parents disregarding children and children disregarding parents) distributions first. 
    - Parent distribution is all heterosexual couples.
    - Correction for gender via multiplying female heights by 1.08.
    - Overplotting is an issue from discretization.

***

```{r galton, results='hide', message=FALSE, warning=FALSE}
library(UsingR); data(galton); library(reshape); long <- melt(galton)
g <- ggplot(long, aes(x = value, fill = variable)) 
g <- g + geom_histogram(colour = "black", binwidth=1) 
g <- g + facet_grid(. ~ variable)
g
```

***

## Finding the middle via least squares

* Consider only the children's heights.
    - How could one describe the "middle"?
    - One definition, let $Y_i$ be the height of child $i$ for $i = 1, \ldots, n = 928$, then define the middle as the value of $\mu$ that minimizes $$\sum_{i=1}^n (Y_i - \mu)^2$$

* This is physical center of mass of the histrogram.
* You might have guessed that the answer $\mu = \bar Y$.

***

## Experiment
##### Use R studio's manipulate to see what value of $\mu$ minimizes the sum of the squared deviations.

```{r manipulate, results='hide', message=FALSE, warning=FALSE}
library(manipulate) ; library(ggplot2)
myHist <- function(mu){
    mse <- mean((galton$child - mu)^2)
    g <- ggplot(galton, aes(x = child)) + geom_histogram(fill = "salmon", colour = "black", binwidth=1)
    g <- g + geom_vline(xintercept = mu, size = 3)
    g <- g + ggtitle(paste("mu = ", mu, ", MSE = ", round(mse, 2), sep = ""))
    g
}
# manipulate(myHist(mu), mu = slider(62, 74, step = 0.5))
```

***

## The least squares est. is the emperical mean

```{r emperical mean}
g <- ggplot(galton, aes(x = child)) + geom_histogram(fill = "salmon", colour = "black", binwidth=1)
g <- g + geom_vline(xintercept = mean(galton$child), size = 3)
g
```

***

## Comparing children's heights and their parents' heights

These points are over-plotted (Next plot shows a better scatterplot of the data).
```{r scatterplot, results='hide', message=FALSE, warning=FALSE}
ggplot(galton, aes(x = parent, y = child)) + geom_point()
```

***

Size of point represents number of points at that (X, Y) combination.
```{r scatterplot2, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")                    
g
```

***

## Regression through the origin

* Suppose that $X_i$ are the parents' heights.
* Consider picking the slope $\beta$ that minimizes $$\sum_{i=1}^n (Y_i - X_i \beta)^2$$
* This is exactly using the origin as a pivot point picking the line that minimizes the sum of the squared vertical distances of the points to the line
* Use R studio's manipulate function to experiment
* Subtract the means so that the origin is the mean of the parent and children's heights

***

```{r manipulate2, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
y <- galton$child - mean(galton$child)
x <- galton$parent - mean(galton$parent)
freqData <- as.data.frame(table(x, y))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
myPlot <- function(beta){
    g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
    g <- g  + scale_size(range = c(2, 20), guide = "none" )
    g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
    g <- g + geom_point(aes(colour=freq, size = freq))
    g <- g + scale_colour_gradient(low = "lightblue", high="white")                     
    g <- g + geom_abline(intercept = 0, slope = beta, size = 3)
    mse <- mean( (y - beta * x) ^2 )
    g <- g + ggtitle(paste("beta = ", beta, "mse = ", round(mse, 3)))
    g
}
# manipulate(myPlot(beta), beta = slider(0.6, 1.2, step = 0.02))
```

***


